---
title: "Histocial CPUE for five flatfish species in the Pacific region"
author: "Sean Anderson and Elise Keppel"
output: html_document
date: "`r format(Sys.Date(), '%B %d, %Y')`"
---

```{r, knitr-opts, echo=FALSE, cache=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.asp = 0.618,
  autodep = TRUE,
  cache = TRUE,
  cache.comments = FALSE
)
```

```{r, message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)
library(gfplot)
```

Extract catch and effort data from all databases for species of interest. This will only work at PBS an authorized computer.

```{r, eval=FALSE}
species <- c("Arrowtooth Flounder", "English Sole", "Dover Sole", 
  "Southern Rock Sole", "Petrale Sole") 
d <- gfplot::get_cpue_historic(species, fishing_year = FALSE, end_year = 2017)
saveRDS(d, file = "data/flatfish-cpue-data.rds", compress = FALSE)
```

Read the data in, clean up the column names, and create columns for the major statistical areas:

```{r}
d <- readRDS("data/flatfish-cpue-data.rds")
d <- rename(d, year = fyear)
d <- mutate(d, species_common_name = tolower(species_common_name))
```

Filter down to remove all rows of data with missing hours fished, missing total catch, or when the total catch or hours fished are 0.

```{r}
d <- d %>%
  filter(!is.na(hours_fished), !is.na(total), total > 0, hours_fished > 0) %>%
  filter(!is.na(year))
```

Now let's calculate the arithmetic and geometric mean CPUE for each species in each year:

```{r}
d_sum <- d %>%
  group_by(area, year, species_common_name) %>%
  summarise(
    catch = sum(total, na.rm = TRUE),
    sum_hours_fished = sum(hours_fished, na.rm = TRUE),
    arith_cpue = sum(total, na.rm = TRUE) / sum(hours_fished, na.rm = TRUE),
    geo_cpue = exp(mean(log(total / hours_fished), na.rm = TRUE))
  ) %>%
  ungroup()
```

Let's create a version that is scaled by the geometric mean for plotting to mimic past research documents:

```{r}
d_scaled <- d_sum %>%
  mutate(arith_cpue = arith_cpue / exp(mean(log(arith_cpue)))) %>%
  mutate(geo_cpue = geo_cpue / exp(mean(log(geo_cpue))))
```

Plot of the scaled version with species in different rows:

```{r}
p <- d_scaled %>%
  tidyr::gather(cpue_type, cpue_value, arith_cpue:geo_cpue) %>%
  ggplot(aes(year, cpue_value, linetype = cpue_type)) +
  geom_line() +
  geom_vline(xintercept = 1996, lty = 3) +
  facet_grid(species_common_name~area, scales = "free_y") +
  xlab("") +
  ylab("CPUE divided by geometric mean") +
  ylim(0, NA) +
  theme_pbs() +
  labs(linetype = "CPUE type")
p
```

# Standardizing data from before 1996

Let's explore possible index standardization.

```{r pre-1996-fits, message=FALSE, warning=FALSE}
d <- mutate(d, cpue = total / hours_fished)
d_pre1996 <- filter(d, year < 1996)
d_pre1996$year_factor <- as.factor(d_pre1996$year)

out_stand <- plyr::dlply(d_pre1996, c("area", "species_common_name"), function(x) {
  
  message(unique(x$area), " - ", unique(x$species_common_name))
  x$locality_factor <- gfplot::f(x$locality_description, ref = get_most_common_level)
  x$month_factor <- gfplot::f(x$month, ref = get_most_common_level)
  
  m <- lm(log(cpue) ~ year_factor + locality_factor + month_factor, data = x)
  
  nd <- data.frame(year_factor = unique(x$year_factor), 
    locality_factor = 
      factor(levels(x$locality_factor)[[1]], levels = levels(x$locality_factor)),
    month_factor = 
      factor(levels(x$month_factor)[[1]], levels = levels(x$month_factor))
  ) %>% 
    arrange(year_factor)
  nd$cpue_pred <- predict(m, newdata = nd)
  nd$cpue_pred_se <- predict(m, newdata = nd, se = TRUE)$se.fit
  list(dat = nd, model = m, spp = unique(x$species_common_name))
})

out_stand_df <- plyr::ldply(out_stand, function(x) x$dat)

out_m <- lapply(out_stand, function(xx) {
  x <- broom::tidy(xx$model, conf.int = TRUE)
  x$term_group <- gsub("([a-z]+)_[a-zA-Z0-9. \\>\\(\\)-\\/']*", "\\1", x$term)
  g <- ggplot(x, aes(term, estimate, ymin = conf.low, ymax = conf.high)) +
    geom_pointrange() +
    facet_wrap(~term_group, scales = "free", ncol = 4) + 
    coord_flip() +
    ggtitle(xx$spp) +
    theme_pbs()
  ggsave(paste0("figs/pre-1996-coefs-", xx$sp, ".pdf"), width = 15, height = 8)
})

pre_1996_cpue <- out_stand_df %>% 
  mutate(year = as.numeric(as.character(year_factor))) %>% 
  left_join(select(filter(d_sum, year < 1996), area, species_common_name, year, geo_cpue), 
    by = c("area", "species_common_name", "year")) %>% 
  as_tibble()
```

Note a huge difference. Blue is standardized, black is the raw geometric mean by year.

```{r, fig.asp=0.3}
pre_1996_cpue %>% group_by(area, species_common_name) %>% 
  mutate(
    geo_cpue = geo_cpue / exp(mean(log(geo_cpue))),
    cpue_pred = log(exp(cpue_pred) / exp(mean(log(exp(cpue_pred))))),
    cpue_stand_lwr = exp(cpue_pred - 2 * cpue_pred_se),
    cpue_stand_upr = exp(cpue_pred + 2 * cpue_pred_se)
  ) %>% 
  ggplot(aes(year, geo_cpue)) +
  geom_line(colour = "grey30") +
  geom_line(aes(y = exp(cpue_pred)), colour = "blue") +
  geom_ribbon(aes(ymin = cpue_stand_lwr, ymax = cpue_stand_upr), fill = "blue", alpha = 0.1) +
  facet_grid(species_common_name~area, scales = "free_y") +
  xlab("") +
  ylab("CPUE") +
  coord_cartesian(ylim = c(0, 8), expand = FALSE) +
  theme_pbs()
```

# Standardizing the data from after 1996

```{r, eval=FALSE}
cpue_dat <- gfplot::get_cpue_index(gear = "bottom trawl", min_year = 1996)
```

```{r, echo=FALSE}
cpue_dat <- readRDS("../gfsynopsis/report/data-cache2/pbs-cpue-index.rds")
```

```{r post-1996-fits, message=FALSE, warning=FALSE, cache=FALSE}
if (!file.exists("data/cpue-post-1996-fits.rds")) {
  post_1996_cpue <- purrr::map_df(unique(d$species_common_name), function(x) {
    message(x)
    o <- gfsynopsis:::fit_cpue_indices(cpue_dat, species = x,
      areas = c("5[CDE]+", "5[AB]+", "3[CD]+"), center = TRUE)
    o$species_common_name <- x
    o
  })
  saveRDS(post_1996_cpue, file = "data/cpue-post-1996-fits.rds", compress = FALSE)
} else {
  post_1996_cpue <- readRDS("data/cpue-post-1996-fits.rds")
}
```

```{r}
ggplot(post_1996_cpue, aes(year, est)) +
  geom_line(colour = "blue") +
  geom_line(aes(y = est_unstandardized), colour = "grey30") +
  geom_ribbon(aes(ymin = lwr, ymax = upr), fill = "blue", alpha = 0.1) +
  facet_grid(species_common_name~area, scales = "free_y") +
  xlab("") +
  facet_grid(species_common_name~area, scales = "free_y") +
  ylab("CPUE") +
  theme_pbs()
```

Save the derived data:

```{r}
raw_summarized_cpue <- d_sum
save(pre_1996_cpue, post_1996_cpue, raw_summarized_cpue, 
  file = "data/flatfish-cpue.rda", compress = TRUE)
```
